Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-05-15 paper/manuscript.Rmd
add Canada to affiliation

---
title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: 
    ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
   University of Newfoundland
    ^2^Department of Biology, Memorial University of Newfoundland, Canada
    #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output:
   github_document: default
   bookdown::word_document2:
     reference_docx: template.docx
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
bibliography: ../references/swim.bib
csl: ecology.csl
---

@@

Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
translocations and introductions throughout the province [@Bergerud1989].
Typical caribou predators such as black bears (*Ursus americanus*) and wolves
(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
(*Canis latrans*), which predate calves but it remains unclear what threat they
pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation
to adults, we propose that forage limitation is more likely to drive movement
between islands than predator avoidance. For several decades after their
introduction to Fogo Island, the caribou population did not exceed ~100
individuals [@Bergerud1989, Newfoundland and Labrador Wildlife Division,
unpublished data]. During the 1990s, population density reached ~300 animals and
anecdotal evidence suggests caribou were first seen on nearby islands during
this same period. Increased competition for resources on Fogo Island due to this
increase in density could have precipitated movement to other islands.

@@

2) At which temporal and spatial scales do caribou choose to swim
between islands? Does the time of day, weather, season, or availability
of food influence swimming?

3) Does reproductive status (presence of a calf-at-heel) or demographic class
(age or sex) influence propensity for oceanic swimming? Our initial observation
was of an adult male, but all collared individuals were adult females. Are
younger caribou more likely to swim if they are unable to compete with older
individuals for resources? Are males or females without a calf-at-heel more
likely to swim in the summer given they are not restricted based on presence of
a calf-at-heel, as females are?

4) If swimming is a function of density-dependent habitat selection, what are
the costs of movement between patches? IFD assumes zero-cost movement, but could
the risk and energetic costs associated with oceanic swimming be estimated using
the potential fitness deficit that caribou are willing to tolerate before
swimming to a new island?

5) In relation to all the previous questions: how do individuals vary in their
propensity to swim and their assessment of the costs and benefits of swimming
between islands? We observed some individuals that regularly engaged in swimming
events, while others were never observed swimming in the ocean? Our existing
research on individual behavioural variation in caribou movement ecology
]@Webber2020] provides a basis to investigate the role of inherent individual
differences in swimming behaviour. Specifically, the forage limitation
hypothesis applies uniformly across populations, but individuals could evaluate
the costs and benefits of swimming differently.

>>>>

---
title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: 
    ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
   University of Newfoundland, Canada
    ^2^Department of Biology, Memorial University of Newfoundland, Canada
    #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output:
   github_document: default
   bookdown::word_document2:
     reference_docx: template.docx
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
bibliography: ../references/swim.bib
csl: ecology.csl
---

@@

Caribou were introduced to Fogo Island between 1964–67 as part of a series of
translocations throughout the province [@Bergerud1989]. Typical caribou
predators such as black bears (*Ursus americanus*) and wolves (*Canis lupus*)
are absent from Fogo Island; the dominant predator is coyote (*Canis latrans*),
which predate calves but it remains unclear what threat they pose, if any, to
adult caribou [@Rayl2014]. Given the minimal risk of predation to adults and low
density of coyote on Fogo Island, we propose that forage limitation is more
likely to drive movement between islands than predator avoidance. For several
decades after their introduction to Fogo Island, the caribou population did not
exceed ~100 individuals [@Bergerud1989, Newfoundland and Labrador Wildlife
Division, unpublished data]. During the 1990s, population density reached ~300
animals and anecdotal evidence suggests caribou were first seen on nearby
islands during this same period. Increased competition for resources on Fogo
Island due to this increase in density could have precipitated movement to other
islands.

@@

<!--2) At which temporal and spatial scales do caribou choose to swim
between islands? Does the time of day, weather, season, or availability
of food influence swimming?-->

<!--3) Does reproductive status (presence of a calf-at-heel) or demographic class
(age or sex) influence propensity for oceanic swimming? Our initial observation
was of an adult male, but all collared individuals were adult females. Are
younger caribou more likely to swim if they are unable to compete with older
individuals for resources? Are males or females without a calf-at-heel more
likely to swim in the summer given they are not restricted based on presence of
a calf-at-heel, as females are? -->

2) If swimming is a function of density-dependent habitat selection, what are
the costs of movement between patches? IFD assumes zero-cost movement, but could
the risk and energetic costs associated with oceanic swimming be estimated using
the potential fitness deficit that caribou are willing to tolerate before
swimming to a new island?

3) In relation to questions 1 and 2: how do individuals vary in their
propensity to swim and their assessment of the costs and benefits of swimming
between islands? We observed some individuals that regularly engaged in swimming
events, while others were never observed swimming in the ocean? Our existing
research on individual behavioural variation in caribou movement ecology
]@Webber2020] provides a basis to investigate the role of inherent individual
differences in swimming behaviour. Specifically, the forage limitation
hypothesis applies uniformly across populations, but individuals could evaluate
the costs and benefits of swimming differently.
